---
title: "Effect of Host Sex effect and Infection Mode on B. pahangi gene expression"
author: Christopher Holt
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# Libraries
```{r}

library(tidyverse)
library(FactoMineR)
library(edgeR)
library(WGCNA)
library(pvclust)
library(devtools)
library(vegan)
library(circlize)
library(ggdendro)
library(variancePartition)


devtools::source_url("https://raw.githubusercontent.com/obigriffith/biostar-tutorials/master/Heatmaps/heatmap.3.R")

```

## Define Functions
```{r}
create_matrix <- function(df, row, centre, column){
  df <- as.data.frame(df)
  test <- df %>% dplyr::select(all_of(row), all_of(centre), all_of(column))
  replicates.test <- unique(test[[column]])
  tmp.matrix <- tibble(gene=unique(df[[row]]))
  for (replicate_uniq in replicates.test){
    tmp <- test %>% filter(!!as.symbol(column) == replicate_uniq)
    colnames(tmp)[2] <- replicate_uniq
    tmp[[column]] <- NULL
    tmp.matrix <- inner_join(tmp.matrix, tmp, by = c('gene' = row))
  }
  
  tmp.matrix <- tmp.matrix %>% tibble::column_to_rownames('gene')
  return(tmp.matrix)
  
}

calculate_TPM <- function(counts.mat, counts.orig){
  gene_length <- counts.orig %>% dplyr::select('Name', 'length') %>% unique
  for (i in 1:ncol(counts.mat)){
    counts.mat[,i] <- counts.mat[,i]/gene_length[,2]
    counts.mat[,i] <- counts.mat[,i]/(sum(counts.mat[,i])/1000000)
  }
  
  return(counts.mat)
}

factor_groups <- function(GROUPS){
  
  for (col in colnames(GROUPS)){
    GROUPS[[col]] = factor(GROUPS[[col]], levels = unique(GROUPS[[col]]))
  }
  
  return(GROUPS)
}


find_soft_power <- function(sft){
  df <- as.data.frame(cbind(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2]))
  y <- -sign(sft$fitIndices[,3])*sft$fitIndices[,2]
  dy <- diff(y) 
  softpower <- which(abs(diff(y)) < 0.05)[1]
  return(softpower)
}

eigengene_invert_id <- function(tpm.de.wgcna, mergedColors, mergedMEs){
  tpm.de <- tpm.de.wgcna
  tpm.de.wgcna$invert <- T
  tpm.de.wgcna$module <- mergedColors
  for(i in 1:nrow(tpm.de.wgcna)){
    if(cor(t(tpm.de[i,]), mergedMEs[,which(colnames(mergedMEs) == paste0("ME",tpm.de.wgcna$module[i]))], method = "pearson") >= 0){
      tpm.de.wgcna$invert[i] <- F
    }
  }
  return(tpm.de.wgcna)
}

wgcna_heatmap_reorder <- function(tpm.de.wgcna){
  clusters <- as.data.frame(table(tpm.de.wgcna$module))
  clusters <- clusters[order(-clusters[,2]),1]

  tpm.de.wgcna.reordered <- as.data.frame(matrix(nrow = 0,
                                                 ncol = ncol(tpm.de.wgcna)))
  for(i in 1:length(clusters)){
    tpm.de.wgcna.reordered <- as.data.frame(rbind(tpm.de.wgcna.reordered,
                                                  tpm.de.wgcna[tpm.de.wgcna$module == clusters[i] & tpm.de.wgcna$invert == F,],
                                                  tpm.de.wgcna[tpm.de.wgcna$module == clusters[i] & tpm.de.wgcna$invert == T,]))
  }
  return(tpm.de.wgcna.reordered)
}

get_heatmap_separators <- function(vector){
  sep <- c()
  for(i in 2:length(unique(vector))){
    sep[length(sep) + 1] <- min(which(vector == unique(vector)[i])) - 1
  }
  return(sep)
}


eigengene_refit <- function(tpm.de, mergedMEs, min_pearson_corr = 0.8){
  tpm.de.wgcna <- tpm.de
  tpm.de.wgcna$invert <- T
  tpm.de.wgcna$module <- NA
  for(i in 1:nrow(tpm.de)){
    minimum_correlation <- min_pearson_corr
    for(j in 1:ncol(mergedMEs)){
      cor <- abs(cor(t(tpm.de[i,]), mergedMEs[,j], method = "pearson"))
      if(cor > minimum_correlation){
        minimum_correlation <- cor
        tpm.de.wgcna$module[i] <- colnames(mergedMEs)[j]
      }
    }
    if(!is.na(tpm.de.wgcna$module[i])){
      if(cor(t(tpm.de[i,]), mergedMEs[,which(colnames(mergedMEs) == tpm.de.wgcna$module[i])], method = "pearson") > 0){
        tpm.de.wgcna$invert[i] <- F
      }
    }else{
      tpm.de.wgcna$invert[i] <- NA
    }

  }
  tpm.de.wgcna$module <- gsub("^ME","",tpm.de.wgcna$module)
  return(tpm.de.wgcna)
}


consolidate_functional_terms <- function(geneinfo){
  geneinfo_genes <- unique(geneinfo$gene)
  ipr <- data.frame()
  gobio <- data.frame()
  gocell <- data.frame()
  gomol <- data.frame()

  ipr.df <- as.data.frame(unlist(strsplit(paste(geneinfo$interpro_description, collapse = "|"),  split = "[|]")))
  gobio.df <- as.data.frame(unlist(strsplit(paste(geneinfo$go_biologicalprocess, collapse = "|"),  split = "[|]")))
  gocell.df <- as.data.frame(unlist(strsplit(paste(geneinfo$go_cellularcomponent, collapse = "|"),  split = "[|]")))
  gomol.df <- as.data.frame(unlist(strsplit(paste(geneinfo$go_molecularfunction, collapse = "|"),  split = "[|]")))

  ipr <- rbind(ipr, ipr.df)
  gobio <- rbind(gobio, gobio.df)
  gocell <- rbind(gocell, gocell.df)
  gomol <- rbind(gomol, gomol.df)

  ipr <- as.data.frame(table(ipr))
  gobio <- as.data.frame(table(gobio))
  gocell <- as.data.frame(table(gocell))
  gomol <- as.data.frame(table(gomol))

  functionalterms.list <- list(ipr = ipr, gobio = gobio, gocell = gocell, gomol = gomol)

  return(functionalterms.list)
}



functionaltermenrichment <- function(genes, geneinfo, functionalterms.list){

  geneinfo.subset <- geneinfo[geneinfo$gene %in% genes,]
  term <- c()
  clusteroccurences <- c()
  genomeoccurences <- c()
  pvalue <- c()
  correctedpvalue <- c()
  oddsratio <- c()
  clusterProportion <- c()
  genomeProportion <- c()
  clusterSize <- c()
  genomeSize <- c()
  newOddsRatio <- c()

  functionalterms.list.subset <- consolidate_functional_terms(geneinfo.subset)

  for(i in 1:length(functionalterms.list)){
    for(j in 1:nrow(functionalterms.list[[i]])){

      freq.all <- functionalterms.list[[i]][j,2] ## How amany times the IPR/GO term appears in the entire geneinfo file

      ## determine if the functional term is in the subset
      ## If it isn't, then the frequency is set to 0
      ## If it is in the subset, then extract the frequency of the IPR/GO term from within the subset
      freq.subset <- ifelse(functionalterms.list[[i]][j,1] %in% functionalterms.list.subset[[i]][,1],
                            functionalterms.list.subset[[i]][functionalterms.list.subset[[i]][,1] == as.character(functionalterms.list[[i]][j,1]),2],
                            0)

      ## Number of genes in the whole genome
      genes.all <- nrow(geneinfo)

      ## Number of genes in the cluster
      genes.subset <- nrow(geneinfo.subset)
      # [col.row]
      ## [1,1] is frequency is the subset, [1,2] is remaining frequency in rest of gene term file (col 1 total is total frequency of term in genome)
      ## [2,1] is no. genes that do not have this term in the subset, [2,2] is no. genes that do not have this term in genome (minus genes in subset)
      fisherexact.matrix <- matrix(c(freq.subset, freq.all - freq.subset,
                                     genes.subset - freq.subset, genes.all - genes.subset - freq.all + freq.subset),
                                   nrow = 2,
                                   ncol = 2)
      fisher.test <- fisher.test(fisherexact.matrix)

      term[length(term) + 1] <- as.character(functionalterms.list[[i]][j,1])
      clusteroccurences[length(clusteroccurences) + 1] <- as.numeric(as.character(freq.subset))
      genomeoccurences[length(genomeoccurences) + 1] <- as.numeric(as.character(freq.all))
      pvalue[length(pvalue) + 1] <- as.numeric(as.character(fisher.test$p.value))
      correctedpvalue[length(correctedpvalue) + 1] <- p.adjust(as.numeric(as.character(fisher.test$p.value)), method = "fdr", n = nrow(functionalterms.list[[i]]))
      oddsratio[length(oddsratio) + 1] <- as.numeric(as.character(fisher.test$estimate)) ## if < 1, term is under represented in the clutser

      clusterProportion[length(clusterProportion) + 1] <- as.numeric(fisherexact.matrix[1,1]/fisherexact.matrix[2,1])
      clusterSize[length(clusterSize) + 1] <- as.numeric(fisherexact.matrix[2,1])
      genomeProportion[length(genomeProportion) + 1] <- as.numeric(fisherexact.matrix[1,2]/fisherexact.matrix[2,2])
      genomeSize[length(genomeSize) + 1] <- as.numeric(fisherexact.matrix[2,2])

      newOddsRatio[length(newOddsRatio) + 1] <- as.numeric((fisherexact.matrix[1,1]/fisherexact.matrix[2,1])/(fisherexact.matrix[1,2]/fisherexact.matrix[2,2]))

    }
  }

  terms.df <- as.data.frame(cbind(term,
                                  clusteroccurences,
                                  genomeoccurences,
                                  pvalue,
                                  correctedpvalue,
                                  oddsratio,
                                  clusterProportion,
                                  clusterSize,
                                  genomeProportion,
                                  genomeSize,
                                  newOddsRatio))
  terms.df <- terms.df[order(as.numeric(as.character(terms.df$pvalue))),]
  return(terms.df)
}



variance_function <- function(counts.to.use, groups.to.use, form){
  y <- edgeR::DGEList(counts = counts.to.use)
  y <- edgeR::calcNormFactors(y)
  design <- stats::model.matrix(~groups.to.use$Infection.JirdSex.WormSex)
  
  y <- limma::voom(y, design)
  
  varPart <- fitExtractVarPartModel(y, form, groups.to.use)
  
  varPart.df <- gg_plotvar(varPart)
  
  return(varPart.df)
}

gg_plotvar <- function(varPart){
  varPart.df <- data.frame()
  for(col in colnames(varPart)){
    df <- as.data.frame(varPart) %>% select(all_of(col))
    df$attribute <- col
    colnames(df)[1] <- 'Variance_Explained'
    
    varPart.df <- rbind(varPart.df, df)
  }
  
  
  varPart.df$attribute <- factor(varPart.df$attribute, levels = unique(varPart.df$attribute))
  
  print(
    ggplot(varPart.df, aes(x = attribute, y = Variance_Explained * 100)) + geom_boxplot() + labs(y = 'Variance explained (%)')
  )
  
  return(varPart.df)

}


ggplot_legend <- function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)

}


median_variance <- function(varPart.df, attr_list){
  
  for(attr in attr_list){
    message(paste0(attr))
      print(
    median(varPart.df[varPart.df$attribute == paste0(attr),]$Variance_Explained)
  )
  }
}


```



# Set up directories

```{r Directories}
# Set up Directories
cwd <- getwd()
OUTPUT_DIR <- paste0(cwd, "/output_files/")
FILE_DIR <- paste0(OUTPUT_DIR, "/files/")
FIGURE_DIR <- paste0(OUTPUT_DIR, '/figures/')

dir.create(OUTPUT_DIR, showWarnings = F, recursive = T)
dir.create(FILE_DIR, showWarnings = F, recursive = T)
dir.create(FIGURE_DIR, showWarnings = F, recursive = T)
```


## File URLS
```{r}
COUNTS_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/combined.final.counts"
GROUPS_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/groups.csv"
GENEINFO_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/bpahangi_gene.info"
GFF_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/GCA_012070555.1_ASM1207055v1_genomic.gff"
```

## Read in Files
```{r}
## Counts file
counts.orig <- readr::read_delim(COUNTS_URL,
                          "\t", escape_double = FALSE, col_names = FALSE, 
                          trim_ws = TRUE, show_col_types = F)

## Read in groups file
groups <- readr::read_delim(GROUPS_URL,
                          ",", escape_double = FALSE, col_names = FALSE, 
                          trim_ws = TRUE, show_col_types = F)


geneinfo <- readr::read_delim(GENEINFO_URL,
                              delim = '\t', show_col_types = F, col_names = T)


b_pahangi_annotation <- readr::read_delim(GFF_URL,
                            "\t", escape_double = FALSE, col_names = FALSE, 
                            trim_ws = TRUE,
                            show_col_types = F, comment = "#")

```

# Format Data Files

## Format groups file

```{r}
groups <- as.data.frame(groups)

## Male Jirds are Orange, Female Jirds are DarkGreen
groups$X2 <- NA
groups <- groups %>% mutate(X2 = ifelse(grepl("^M",groups$X1), 'orange',
                                        ifelse(grepl('^F', groups$X1), 'darkgreen','white')))

## Male Worms are orange, Female Worms are Blue, MF are white
groups$X3 <- NA
groups <- groups %>% mutate(X3 = ifelse(grepl("-males",groups$X1), 'red',
                                        ifelse(grepl('-females', groups$X1), 'blue','white')))

## IP Infections are Grey, SQ Infections are black
groups$X4 <- NA
groups <- groups %>% mutate(X4 = ifelse(grepl("IP", groups$X1), 'grey',
                                        ifelse(grepl('SQ', groups$X1), 'black', 'white')))

## Add litter to each sample
groups$X5 <- NA
groups <- groups %>% mutate(X5 = ifelse(grepl("M200_IP_Bp|F006_IP_Bp", groups$X1), 'litter_05',
                                        ifelse(grepl("F009_IP_Bp|M204_IP_Bp", groups$X1), 'litter_02',
                                               ifelse(grepl("F012_IP_Bp|M206_IP_Bp", groups$X1), 'litter_03',
                                                      ifelse(grepl("F024_SQ_Bp|M055_SQ_Bp", groups$X1), 'litter_12',
                                                             ifelse(grepl("M057_SQ_Bp", groups$X1), 'litter_13',
                                                                    ifelse(grepl("F015_SQ_Bp", groups$X1), 'litter_08',
                                                                           ifelse(grepl("M069_SQ_Bp", groups$X1), 'CRL',
                                                                                  ifelse(grepl("F022_SQ_Bp", groups$X1), 'litter_11', NA)))))))))

## Add Infection classifier to be able to select and group IP and SQ samples separately
groups$X6 <- NA
groups <- groups %>% mutate(X6 = ifelse(grepl("IP", groups$X1), "IP", "SQ")) 

## Add Host Sex classifier to be able to select and group Male Jird and Female Jird separately
groups$X7 <- NA
groups <- groups %>% mutate(X7 = ifelse(grepl("^F", groups$X1), "FemaleGerbil", "MaleGerbil"))

## Add Worm Sex classifier to be able to select and group Male Worm and Female Worm separately

groups$X8 <- NA
groups <- groups %>% mutate(X8 = ifelse(grepl("-females", groups$X1), "FemaleWorm", 
                                        ifelse(grepl("-males", groups$X1), "MaleWorm", "mf")))

## Combine Infection and Host Sex classifiers 
groups$X9 <- NA
groups$X9 <- paste0(groups$X6, "_", groups$X7)
## Combine Infection and Worm Sex Classifiers
groups$X10 <- NA
groups$X10 <- paste0(groups$X6, "_",  groups$X8)
## Combine Host Sex and Worm Sex classifiers
groups$X11 <- NA
groups$X11 <- paste0(groups$X7, "_", groups$X8)
## Combine all three classifiers together
groups$X12 <- NA
groups$X12 <- paste0(groups$X6, "_", groups$X7, "_", groups$X8)

groups$X13 <- NA
groups <- groups %>% mutate(X13 = ifelse(grepl("IP_FemaleWorm", groups$X10), 'blue',
                                             ifelse(grepl("IP_MaleWorm", groups$X10), 'red',
                                                    ifelse(grepl("SQ_FemaleWorm", groups$X10), 'lightblue',
                                                           ifelse(grepl("SQ_MaleWorm", groups$X10), 'pink',
                                                                  ifelse(grepl("IP_mf", groups$X10), 'darkgreen',
                                                                         ifelse(grepl("SQ_mf", groups$X10), 'lightgreen', NA)))))))



## Set Litter Colour
groups$X14 <- NA
groups <- groups %>% mutate(X14 = ifelse(grepl('litter_05', groups$X5), 'green',
                                        ifelse(grepl('litter_02', groups$X5), 'orange',
                                               ifelse(grepl('litter_03', groups$X5), 'blue',
                                                      ifelse(grepl('litter_12', groups$X5), 'violet',
                                                             ifelse(grepl('litter_13', groups$X5), 'black',
                                                                    ifelse(grepl('litter_08', groups$X5), 'grey',
                                                                           ifelse(grepl('CRL', groups$X5), 'white',
                                                                                  ifelse(grepl('litter_11', groups$X5), 'gold', NA)))))))))
  
groups$X15 <- NA
groups <- groups %>% 
  dplyr::mutate(X15 = ifelse(grepl("F015_SQ_Bp-females", groups$X1), 'blue',
                                           ifelse(grepl("F015_SQ_Bp-males", groups$X1), 'orange',
                                                 ifelse(grepl("F024_SQ_Bp-females", groups$X1), 'green',
                                                        ifelse(grepl('M069_SQ_Bp-males', groups$X1), 'black',
                                                               ifelse(grepl('M057_SQ_Bp-males', groups$X1), 'grey',
                                                                      'white'))))))
groups$X16 <- NA
groups <- groups %>% 
  dplyr::mutate(X16 = ifelse(grepl("F015_SQ_Bp-females", groups$X1), "F015_SQ_Bp-females",
                                           ifelse(grepl("F015_SQ_Bp-males", groups$X1), 'F015_SQ_Bp-males',
                                                 ifelse(grepl("F024_SQ_Bp-females", groups$X1), 'F024_SQ_Bp-females',
                                                        ifelse(grepl('M069_SQ_Bp-males', groups$X1), 'M069_SQ_Bp-males',
                                                               ifelse(grepl('M057_SQ_Bp-males', groups$X1), 'M057_SQ_Bp-males',
                                                                      'not_outlier'))))))
groups$X17 <- NA
groups <- groups %>%
  dplyr::mutate(X17 = ifelse(grepl('MaleGerbil_FemaleWorm', groups$X11), 'blue',
                             ifelse(grepl('MaleGerbil_MaleWorm', groups$X11), 'red',
                                    ifelse(grepl('MaleGerbil_mf', groups$X11), 'darkgreen',
                                           ifelse(grepl('FemaleGerbil_FemaleWorm', groups$X11), 'lightblue',
                                                  ifelse(grepl('FemaleGerbil_MaleWorm', groups$X11), 'pink',
                                                         ifelse(grepl('FemaleGerbil_mf', groups$X11), 'green', NA)))))))

colnames(groups) <- c("Sample", "JirdColour", "WormColour", 
                      "InfectionColour", "Litter", "Infection", "JirdSex",
                      "WormSex", "Infection.JirdSex", "Infection.WormSex", "JirdSex.WormSex", "Infection.JirdSex.WormSex",
                      'Infection.WormSex.Colour', 'LitterColour', 'outlier_samples', 'outliersample_names','JirdSex.WormSex.Colour')

groups <- factor_groups(groups)


```


## Format Annotation File

```{r format annotation file}
b_pahangi_annotation <- b_pahangi_annotation %>% tidyr::separate('X9', into = c("Name", "X9"), sep = ';')
b_pahangi_annotation$X9 <- gsub(".*:", "", b_pahangi_annotation$X9)
b_pahangi_annotation$X9 <- gsub(".*=", "", b_pahangi_annotation$X9)
b_pahangi_annotation$Name <- gsub(".*:", "", b_pahangi_annotation$Name)
b_pahangi_annotation$Name <- gsub(".*=", "", b_pahangi_annotation$Name)

b_pahangi_genes <- b_pahangi_annotation %>% dplyr::filter(X3 == "gene")

b_pahangi_genes$length <- b_pahangi_genes$X5 - b_pahangi_genes$X4

b_pahangi_gene_length <- b_pahangi_genes %>% dplyr::select('Name', 'length')

```

## Format counts table into a matrix

```{r format counts table into a matrix}
colnames(counts.orig) <- c("Name", "counts", "sample")

counts.orig$counts <- as.double(counts.orig$counts)
counts.orig$sample <- str_replace_all(counts.orig$sample, ".counts", "")

counts.orig <- counts.orig %>% dplyr::arrange(sample, Name)
counts.orig <- counts.orig[!grepl("__", counts.orig$Name),]

counts.orig <- dplyr::inner_join(counts.orig, b_pahangi_gene_length, by = "Name") %>% dplyr::arrange(sample, Name)
counts <- create_matrix(df = counts.orig, row = "Name", centre = "counts", column = "sample")

```


## Select genes annotated as mRNA

```{r}
b_pahangi_mRNAs <- b_pahangi_annotation %>% filter(X3 == "mRNA")

b_pahangi_mRNAs <- unique(b_pahangi_mRNAs$X9)

counts <- counts[(rownames(counts) %in% b_pahangi_mRNAs),]

## Select genes in the counts.orig table that are in the counts matrix
counts.orig <- counts.orig[(counts.orig$Name %in% rownames(counts)),]
```

## Calculate TPMs

```{r calculate TPMs}

if(all(rownames(counts) == unique(counts.orig$Name))){ ## Makes sure the genes are in the same order for the correct division
  
  tpm <- calculate_TPM(counts.mat = counts, counts.orig = counts.orig)

}else{
  message('The genes are not in the same order')
}

counts <- counts[,match(groups$Sample,colnames(counts))]
tpm <- tpm[,match(groups$Sample,colnames(tpm))]

```

## Plot rarefaction curve of samples

```{r rarefaction curve, eval=FALSE}
## SFig 2
rarefy.counts <- round(counts,0)
raremax <- round(min(rowSums(t(rarefy.counts))),0)
srare <- vegan::rarefy(t(rarefy.counts),raremax)
rarefy.raw.df <- vegan::rarecurve(t(rarefy.counts), step = round(raremax/10,0), sample = raremax, tidy = T)
rarefy.raw.df <- rarefy.raw.df %>% group_by(Site)

rarefy.df <- as.data.frame(matrix(nrow = 0,
                                  ncol = 5))
rarefy.points.df <- rarefy.df
for(i in unique(rarefy.raw.df$Site)){
  steps <- as.numeric(rarefy.raw.df[rarefy.raw.df$Site == i,]$Sample)
  detected_genes <- as.numeric(rarefy.raw.df[rarefy.raw.df$Site == i,]$Species)
  
  rarefy.df <- as.data.frame(rbind(rarefy.df,
                                   cbind(as.numeric(steps),as.numeric(detected_genes),
                                         as.character(groups[groups$Sample == i,]$Sample),
                                         as.character(groups[groups$Sample == i,]$Infection.WormSex),
                                         as.character(groups[groups$Sample == i,]$Infection.WormSex.Colour))))
  
  rarefy.points.df <- as.data.frame(rbind(rarefy.points.df,
                                          cbind(as.numeric(max(steps)),as.numeric(max(detected_genes)),
                                                as.character(groups[groups$Sample == i,]$Sample),
                                                as.character(groups[groups$Sample == i,]$Infection.WormSex),
                                                as.character(groups[groups$Sample == i,]$Infection.WormSex.Colour))))
  
  
}

col <- as.character(groups$Infection.WormSex.Colour)
names(col) <- as.character(groups$Infection.WormSex)

pdf(paste0(RAREFACTION_DIR, "/rarefaction_curve.pdf"))
  print(
    ggplot()+
      geom_line(mapping=aes(x = as.numeric(as.character(rarefy.df$V1)), 
                            y = as.numeric(as.character(rarefy.df$V2)),
                            group=rarefy.df$V3,color = rarefy.df$V4)) +
      geom_point(mapping=aes(x = as.numeric(as.character(rarefy.points.df$V1)), 
                             y = as.numeric(as.character(rarefy.points.df$V2)),
                             group = rarefy.points.df$V3, color = rarefy.points.df$V4),size = 3) +
      guides(shape = 'none') +
     scale_color_manual(values = col)+
    labs(x="number of reads mapping to genes annotated as mRNA", y="number genes detected", color = "InfectionMode_WormSex") +
    theme_bw() + scale_y_continuous(expand = expansion(mult = c(0.2, .2)))

  )

dev.off()


svg(paste0(RAREFACTION_DIR, "/rarefaction_curve.svg"))
  print(
    ggplot()+
      geom_line(mapping=aes(x = as.numeric(as.character(rarefy.df$V1)), 
                            y = as.numeric(as.character(rarefy.df$V2)),
                            group=rarefy.df$V3,color = rarefy.df$V4)) +
      geom_point(mapping=aes(x = as.numeric(as.character(rarefy.points.df$V1)), 
                             y = as.numeric(as.character(rarefy.points.df$V2)),
                             group = rarefy.points.df$V3, color = rarefy.points.df$V4),size = 3) +
      guides(shape = 'none') +
     scale_color_manual(values = col)+
    labs(x="number of reads mapping to genes annotated as mRNA", y="number genes detected", color = "InfectionMode_WormSex") +
    theme_bw() + scale_y_continuous(expand = expansion(mult = c(0.2, .2)))

  )

dev.off()

jpeg(paste0(RAREFACTION_DIR, "/rarefaction_curve.jpg"))
  print(
    ggplot()+
      geom_line(mapping=aes(x = as.numeric(as.character(rarefy.df$V1)), 
                            y = as.numeric(as.character(rarefy.df$V2)),
                            group=rarefy.df$V3,color = rarefy.df$V4)) +
      geom_point(mapping=aes(x = as.numeric(as.character(rarefy.points.df$V1)), 
                             y = as.numeric(as.character(rarefy.points.df$V2)),
                             group = rarefy.points.df$V3, color = rarefy.points.df$V4),size = 3) +
      guides(shape = 'none') +
     scale_color_manual(values = col)+
    labs(x="number of reads mapping to genes annotated as mRNA", y="number genes detected", color = "InfectionMode_WormSex") +
    theme_bw() + scale_y_continuous(expand = expansion(mult = c(0.2, .2)))

  )

dev.off()

tiff(paste0(RAREFACTION_DIR, "/rarefaction_curve.tif"))
  print(
    ggplot()+
      geom_line(mapping=aes(x = as.numeric(as.character(rarefy.df$V1)), 
                            y = as.numeric(as.character(rarefy.df$V2)),
                            group=rarefy.df$V3,color = rarefy.df$V4)) +
      geom_point(mapping=aes(x = as.numeric(as.character(rarefy.points.df$V1)), 
                             y = as.numeric(as.character(rarefy.points.df$V2)),
                             group = rarefy.points.df$V3, color = rarefy.points.df$V4),size = 3) +
      guides(shape = 'none') +
     scale_color_manual(values = col)+
    labs(x="number of reads mapping to genes annotated as mRNA", y="number genes detected", color = "InfectionMode_WormSex") +
    theme_bw() + scale_y_continuous(expand = expansion(mult = c(0.2, .2)))

  )

dev.off()
```


## Remove SQ Microfilariae
```{r}
## Remove all microfilariae samples and the BpSQ69_male samples
groups <- groups %>% filter(!grepl('SQ_Bp-mf', groups$Sample))
groups <- factor_groups(groups)

counts <- counts[match(rownames(tpm), rownames(counts)),match(groups$Sample, colnames(counts))]
tpm <- tpm[,match(groups$Sample,colnames(tpm))]

```

## Plot Variance
```{r}
## Fig 4

## SQ worms
groups.to.use <- groups %>% filter(grepl("SQ", groups$Sample))
groups.to.use <- factor_groups(groups.to.use)
counts.to.use <- counts[,colnames(counts) %in% groups.to.use$Sample]
form <- ~ (1|Litter) + (1|WormSex) + (1|JirdSex)
varPart.df.sq <- variance_function(counts.to.use, groups.to.use, form)


median_variance(varPart.df.sq, c('Residuals', all.vars(form)))

pdf(paste0(FIGURE_DIR, "/sq_samples_variancePartition_histogram.pdf"),
     width = 10,
     height = 10)
print(
 ggplot(varPart.df.sq, aes(x = Variance_Explained)) + geom_histogram() + facet_wrap(~attribute, ncol = 1) + 
   labs(x = 'Variance explained ', title = 'Subcutaneous Worms') + theme_bw()
)
dev.off()

pdf(paste0(FIGURE_DIR, "/sq_samples_variancePartition.pdf"),
     width = 10,
     height = 10)
print(
  ggplot(varPart.df.sq, aes(x = attribute, y = Variance_Explained * 100)) + 
  geom_boxplot() + labs(y = 'Variance explained (%)', title = 'Subcutaneous Worms') + theme_bw()
)
dev.off()


## IP Worms, no microfilariae
groups.to.use <- groups %>% filter(grepl("IP", groups$Sample)) %>% filter(!(grepl("mf", .$Sample)))
groups.to.use <- factor_groups(groups.to.use)
counts.to.use <- counts[,colnames(counts) %in% groups.to.use$Sample]
form <- ~ (1|Litter) + (1|WormSex) + (1|JirdSex)
varPart.df.ip <- variance_function(counts.to.use, groups.to.use, form)

median_variance(varPart.df.ip, c('Residuals', all.vars(form)))



pdf(paste0(FIGURE_DIR, "/ip_no_mf_samples_variancePartition_histogram.pdf"),
     width = 10,
     height = 10)
print(
 ggplot(varPart.df.ip, aes(x = Variance_Explained)) + geom_histogram() + facet_wrap(~attribute, ncol = 1) + 
   labs(x = 'Variance explained ', title = 'Intraperitoneal Worms') + theme_bw()
)
dev.off()

pdf(paste0(FIGURE_DIR, "/ip_no_mf_samples_variancePartition.pdf"),
     width = 10,
     height = 10)
print(
  ggplot(varPart.df.ip, aes(x = attribute, y = Variance_Explained * 100)) + 
  geom_boxplot() + labs(y = 'Variance explained (%)', title = 'Intraperitoneal Worms') + theme_bw()
)
dev.off()




```


## Identify DE Genes

```{r}

FDRcutoff <- 0.05
cpm.cutoff <- 5/min(colSums(counts)) * 1000000 
print(paste0(nrow(counts)," total genes"))
y <- edgeR::DGEList(counts = counts, group = groups$Infection.JirdSex.WormSex)
y <- edgeR::calcNormFactors(y)
keep <- rowSums(cpm(y) >= cpm.cutoff) >= min(table(groups$Infection.JirdSex.WormSex)) 
keep.df <- as.data.frame(table(keep))
print(paste0(keep.df[keep.df[,1] == F,2]," genes excluded using edgeR filter"))
y <- y[keep, , keep.lib.sizes = F] 
design <- stats::model.matrix(~groups$Infection.JirdSex.WormSex)
y <- estimateDisp(y , design)
fit <- edgeR::glmQLFit(y, design) 
qlf <- edgeR::glmQLFTest(fit, coef = 2:ncol(fit)) 
FDR <- as.data.frame(p.adjust(qlf$table$PValue, method="BH"))
rownames(FDR) <- rownames(qlf$table)
colnames(FDR) <- FDR
degenes <- rownames(FDR[FDR[,1] < FDRcutoff, , drop = F])
print(paste0(length(degenes)," DE genes"))

tpm.de <- tpm[rownames(tpm) %in% degenes,]

```


## Clustering of all the samples with bootstrap support
## Z-score dendrogram
```{r}

## The dendrogram is part of Fig 3
tpm_dendrogram <- pvclust::pvclust(as.data.frame(t(scale(t(log2(tpm.de+1))))), method.dist="cor", method.hclust="average",
                                   nboot=1000, parallel=TRUE)
pdf(paste0(FIGURE_DIR, "/dendrogram.pdf"),
    width = 20,
    height = 10)
  plot(tpm_dendrogram)
invisible(dev.off())


dendro.data <- ggdendro::dendro_data(tpm_dendrogram$hclust)
dendro_order <- dendro.data$labels %>% dplyr::select('label')

sample_order <- factor(as.character(dendro_order$label), levels = as.character(dendro_order$label))
groups <- factor_groups(groups[order(match(groups$Sample, as.character(sample_order))),])

rownames(groups) <- NULL

```

## Clustering of all the samples with bootstrap support

## Plot Principle Component Analysis of the Data
```{r}


## Fig 2
tpm.de <- tpm.de[,as.character(sample_order)]
pca <- FactoMineR::PCA(as.data.frame(t(as.data.frame(t(scale(t(log2(tpm.de+1))))))), graph = FALSE, ncp = ncol(counts) - 1)

pdf(paste0(FIGURE_DIR, "/PCA_of_samples_de_jirdsex.pdf"),
     width = 10,
     height = 10)
  plot(
    ggplot()+
      geom_point(aes(x=pca$ind$coord[,1], y=pca$ind$coord[,2], shape = groups$JirdSex), size = 10) +
      labs(col = "GerbilSex", 
           x = paste0("PC1 (", round(pca$eig[1,2],1), "%)"), 
           y = paste0("PC2 (", round(pca$eig[2,2],1), "%)"))+
      guides(colour = guide_legend(ncol = 2))+
      scale_shape_manual(breaks = c("MaleGerbil", "FemaleGerbil"), values = c(16, 1)) +
      guides(shape = guide_legend("HostSex")) +
      theme_bw()
  )
dev.off()

pdf(paste0(FIGURE_DIR, "/PCA_of_samples_de_infectionmode.pdf"),
     width = 10,
     height = 10)
  plot(
    ggplot()+
      geom_point(aes(x=pca$ind$coord[,1], y=pca$ind$coord[,2], shape = groups$Infection), size = 10) +
      labs(col = "InfectionMode", 
           x = paste0("PC1 (", round(pca$eig[1,2],1), "%)"), 
           y = paste0("PC2 (", round(pca$eig[2,2],1), "%)"))+
      guides(colour = guide_legend(ncol = 2))+
        scale_shape_manual(breaks = c("SQ", "IP"),
                           values = c(16, 1)) +
        guides(shape = guide_legend("infection")) +
      theme_bw()
  )
  
dev.off()



pdf(paste0(FIGURE_DIR, "/PCA_of_samples_de_life_stage.pdf"),
     width = 10,
     height = 10)
  plot(
    ggplot()+
      geom_point(aes(x=pca$ind$coord[,1], y=pca$ind$coord[,2], shape = groups$WormSex), size = 10) +
      labs(col = "LifeStage", 
           x = paste0("PC1 (", round(pca$eig[1,2],1), "%)"), 
           y = paste0("PC2 (", round(pca$eig[2,2],1), "%)"))+
      guides(colour = guide_legend(ncol = 2))+
        scale_shape_manual(breaks = c("MaleWorm", "FemaleWorm", "mf"),
                           values = c(16, 1, 2)) +
        guides(shape = guide_legend("LifeStage")) +
      theme_bw()
  )
dev.off()

```

## WGCNA Clustering
```{r}
wgcna <- as.data.frame(t(tpm.de))
powers <- c(c(1:10), seq(from = 12, to=20, by=2))
sft <- pickSoftThreshold(wgcna, powerVector = powers, verbose = 5)
softpower <- find_soft_power(sft)

adjacency <- adjacency(wgcna, power = softpower)
TOM <- TOMsimilarity(adjacency)
dissTOM <- 1-TOM
geneTree <- hclust(as.dist(dissTOM), method = "average");
minModuleSize <- 1
dynamicMods <- cutreeDynamic(dendro = geneTree, distM = dissTOM,
                             deepSplit = 2, pamRespectsDendro = FALSE,
                             minClusterSize = minModuleSize)
dynamicColors = labels2colors(dynamicMods)
MEList = moduleEigengenes(wgcna, colors = dynamicColors)
MEs = MEList$eigengenes
MEDiss = 1-cor(MEs, use = "pairwise.complete.obs")
METree = hclust(as.dist(MEDiss), method = "average")
MEDissThres = 0.25


merge = mergeCloseModules(wgcna, dynamicColors, cutHeight = MEDissThres, verbose = 3)
mergedColors = merge$colors
mergedMEs = merge$newMEs

tpm.de.wgcna <- eigengene_invert_id(tpm.de, mergedColors, mergedMEs)
tpm.de.wgcna <- wgcna_heatmap_reorder(tpm.de.wgcna)

log2tpm.de <- log2(tpm.de.wgcna[,1:(ncol(tpm.de.wgcna) - 2)] + 1)
zscore.log2tpm.de <- as.data.frame(t(scale(t(log2tpm.de))))
zscore.log2tpm.de <- zscore.log2tpm.de[,as.character(sample_order)]


##-----------------------------------------------------------------------------------------------------------
## Add invert and module column to the zscore dataframe
##-----------------------------------------------------------------------------------------------------------

zscore.log2tpm.de <- inner_join(zscore.log2tpm.de %>% rownames_to_column('gene'), 
           tpm.de.wgcna %>% dplyr::select('invert', 'module') %>% rownames_to_column('gene'), by = 'gene') %>% column_to_rownames('gene')
log2tpm.de <- inner_join(log2tpm.de %>% rownames_to_column('gene'), 
           tpm.de.wgcna %>% dplyr::select('invert', 'module') %>% rownames_to_column('gene'), by = 'gene') %>% column_to_rownames('gene')


hmcol <- colorRampPalette(c("navyblue","white","firebrick3"))(12)
rowcol <- tpm.de.wgcna$module
colcol <- as.character(groups$InfectionColour)
rowsep <- get_heatmap_separators(rowcol)
#colsep <- get_heatmap_separators(colcol)
breaks <- seq(-3,3,by=.5)

colcol1 <- as.character(groups$WormColour)
colcol2 <- as.character(groups$JirdColour)

colcol3 <- cbind(colcol, colcol1, colcol2)

rowcol1 <- zscore.log2tpm.de$invert
rowcol1[rowcol1 == F] <- "grey" ## matching pattern
rowcol1[rowcol1 == T] <- "black" ## inverted pattern

rowcol2 <- t(cbind(rowcol, rowcol1))


##-----------------------------------------------------------------------------------------------------------
## Set the row colours and coloumn colours as matrices
##-----------------------------------------------------------------------------------------------------------

rownames(rowcol2) <- c('Main', 'Inverse')
rowcol2 <- as.matrix((rowcol2))

colnames(colcol3) <- c('Infection', 'WormStage', 'GerbilSex')
colcol3 <- as.matrix(colcol3)



heatmap_matrix <- as.matrix(zscore.log2tpm.de[,(1:(ncol(zscore.log2tpm.de)-2))])
heatmap_matrix <- heatmap_matrix[,as.character(sample_order)]

## Fig 3
pdf(paste0(FIGURE_DIR, "/WGCNA_de_genes.pdf"),
    width = 20,
    height = 10)
heatmap.3(heatmap_matrix,
              col=hmcol,
              trace="none",
              labRow = NA,
              labCol = F,
              Rowv = F,
              Colv = F,
              RowSideColors=rowcol2,
              RowSideColorsSize = 1,
              ColSideColorsSize = 4,
              ColSideColors=colcol3,
              lhei = c(2,8,1),
              breaks = breaks,
              density.info = 'histogram',
              dendrogram = "none")
invisible(dev.off())

```


## Log Version
```{r}

log2tpm.de <- log2tpm.de[,as.character(sample_order)]
heatmap_matrix <- as.matrix(log2tpm.de)
col_fun = circlize::colorRamp2(c(0, 7, 14), c("white", "firebrick1", "firebrick3"))
breaks <- seq(min(heatmap_matrix),max(heatmap_matrix),by=.5)
hmcol.log <- colorRampPalette(c("white","firebrick1","firebrick3"))(length(breaks)-1)



colcol <- as.character(groups$InfectionColour)
colcol1 <- as.character(groups$WormColour)
colcol2 <- as.character(groups$JirdColour)
colcol3 <- cbind(colcol, colcol1, colcol2)
colnames(colcol3) <- c('Infection', 'WormStage', 'GerbilSex')
colcol3 <- as.matrix(colcol3)



heatmap_matrix <- as.matrix(log2tpm.de)

## S3 Fig
pdf(paste0(FIGURE_DIR, "/WGCNA_log_de_genes.pdf"),
    width = 20,
    height = 10)
heatmap.3(heatmap_matrix,
              col=hmcol.log,
              trace="none",
              labRow = NA,
              labCol = F,
              Rowv = T,
              Colv = F,
              RowSideColorsSize = 0.5,
              ColSideColorsSize = 4,
              ColSideColors=colcol3,
              lhei = c(2,8,1),
              breaks = breaks,
              density.info = 'histogram',
              dendrogram = "row")
invisible(dev.off())




```

## Number of genes in each cluster
```{r}
table_of_gene_number_in_heatmap <- tibble('Type_of_Comparison' = as.character() ,
                                          'Number_of_Genes_in_Heatmap' = as.numeric())

for(cluster in unique(tpm.de.wgcna$module)){
  for(invert_type in unique(tpm.de.wgcna$invert)){
      table_of_gene_number_in_heatmap <- table_of_gene_number_in_heatmap %>% 
        add_row(Type_of_Comparison = paste0(cluster, "_", invert_type),
          Number_of_Genes_in_Heatmap = nrow(tpm.de.wgcna %>% filter(module == cluster & invert == invert_type)))
  }
}

```



## Functional Enrichment

```{r}
#### Functional term enrichment
terms.colnames <- c("term","clusteroccurences","genomeoccurences","pvalue","correctedpvalue","oddsratio","clusterProportion",
                    "clusterSize", "genomeProportion", "genomeSize", "newOddsRatio" ,"module","invert")
terms.wgcna <- as.data.frame(matrix(nrow = 0,
                                    ncol = length(terms.colnames)))
colnames(terms.wgcna) <- terms.colnames

functionalterms.list <- consolidate_functional_terms(geneinfo)

for(i in 1:length(unique(tpm.de.wgcna$module))){
  ## First part selects the genes that match the invert value and module ID which becomes the gene list for the functionalenrichment function, then paste in gene info to get the gene descriptions.
  terms.wgcna.f <- as.data.frame(cbind(functionaltermenrichment(rownames(zscore.log2tpm.de)[zscore.log2tpm.de$module == unique(zscore.log2tpm.de$module)[i] & zscore.log2tpm.de$invert == F],geneinfo, functionalterms.list),
                                       unique(zscore.log2tpm.de$module)[i],F))
  terms.wgcna.t <- as.data.frame(cbind(functionaltermenrichment(rownames(zscore.log2tpm.de)[zscore.log2tpm.de$module == unique(zscore.log2tpm.de$module)[i] & zscore.log2tpm.de$invert == T],geneinfo, functionalterms.list),
                                       unique(zscore.log2tpm.de$module)[i],T))
  
  colnames(terms.wgcna.f) <- terms.colnames
  colnames(terms.wgcna.t) <- terms.colnames
  terms.wgcna <- as.data.frame(rbind(terms.wgcna,
                                     terms.wgcna.f, 
                                     terms.wgcna.t))
}

pvalue.cutoff <- 0.05
terms.wgcna.sig <- terms.wgcna[as.numeric(as.character(terms.wgcna$correctedpvalue)) < pvalue.cutoff,]
print(paste0(nrow(terms.wgcna.sig)," significantly over-/under-represented terms in WGCNA modules"))

```

## Write out functional terms
```{r}
## These two files are S1 Table
## Table of number of genes per cluster
write.table(table_of_gene_number_in_heatmap ,
            paste0(FILE_DIR,"/table_of_heatmap.tsv"), 
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")
## Supplementary File 1
write.table(zscore.log2tpm.de %>% rownames_to_column('gene'),
            paste0(FILE_DIR, "/Supplementary_File_1.tsv"),
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")


## All wgcna terms
write.table(terms.wgcna,
            paste0(FILE_DIR,"/functionaltermenrichment_wgcna.tsv"), 
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")

## Supplemental File 4
## All significantly enriched terms with an odds ratio >= 1
write.table(terms.wgcna.sig %>% filter(oddsratio >= 1) ,
            paste0(FILE_DIR,"/functionaltermenrichment_sig_wgcna_overrep.tsv"), 
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")
## All significantly enriched terms with an odds ratio < 1
write.table(terms.wgcna.sig %>% filter(oddsratio < 1) ,
            paste0(FILE_DIR,"/functionaltermenrichment_sig_wgcna_underrep.tsv"), 
            row.names = F,
            col.names = T,
            quote = F,
            sep = "\t")





```





