---
title: "Pairwise Comparisons"
author: Christopher Holt
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# Libraries
```{r}

library(tidyverse)
library(FactoMineR)
library(edgeR)
library(WGCNA)
library(pvclust)
library(devtools)
library(vegan)
library(circlize)
library(ggdendro)
library(variancePartition)


devtools::source_url("https://raw.githubusercontent.com/obigriffith/biostar-tutorials/master/Heatmaps/heatmap.3.R")

```

## Define Functions
```{r}
create_matrix <- function(df, row, centre, column){
  df <- as.data.frame(df)
  test <- df %>% select(all_of(row), all_of(centre), all_of(column))
  replicates.test <- unique(test[[column]])
  tmp.matrix <- tibble(gene=unique(df[[row]]))
  for (replicate_uniq in replicates.test){
    tmp <- test %>% filter(!!as.symbol(column) == replicate_uniq)
    colnames(tmp)[2] <- replicate_uniq
    tmp[[column]] <- NULL
    tmp.matrix <- inner_join(tmp.matrix, tmp, by = c('gene' = row))
  }
  
  tmp.matrix <- tmp.matrix %>% tibble::column_to_rownames('gene')
  return(tmp.matrix)
  
}

calculate_TPM <- function(counts.mat, counts.orig){
  gene_length <- counts.orig %>% select('Name', 'length') %>% unique
  for (i in 1:ncol(counts.mat)){
    counts.mat[,i] <- counts.mat[,i]/gene_length[,2]
    counts.mat[,i] <- counts.mat[,i]/(sum(counts.mat[,i])/1000000)
  }
  
  return(counts.mat)
}

factor_groups <- function(GROUPS){
  
  for (col in colnames(GROUPS)){
    GROUPS[[col]] = factor(GROUPS[[col]], levels = unique(GROUPS[[col]]))
  }
  
  return(GROUPS)
}
fix_row_names <- function(df){
  df <- as.data.frame(df)
  rownames(df) <- df[,1]
  df <- df[,-1]
  
  return(df)
}


factor_groups <- function(GROUPS){
  
  for (col in colnames(GROUPS)){
    GROUPS[[col]] = factor(GROUPS[[col]], levels = unique(GROUPS[[col]]))
  }
  
  return(GROUPS)
}


find_soft_power <- function(sft){
  df <- as.data.frame(cbind(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2]))
  y <- -sign(sft$fitIndices[,3])*sft$fitIndices[,2]
  dy <- diff(y) 
  softpower <- which(abs(diff(y)) < 0.05)[1]
  return(softpower)
}

eigengene_invert_id <- function(tpm.de.wgcna, mergedColors, mergedMEs){
  tpm.de <- tpm.de.wgcna
  tpm.de.wgcna$invert <- T
  tpm.de.wgcna$module <- mergedColors
  for(i in 1:nrow(tpm.de.wgcna)){
    if(cor(t(tpm.de[i,]), mergedMEs[,which(colnames(mergedMEs) == paste0("ME",tpm.de.wgcna$module[i]))], method = "pearson") >= 0){
      tpm.de.wgcna$invert[i] <- F
    }
  }
  return(tpm.de.wgcna)
}

wgcna_heatmap_reorder <- function(tpm.de.wgcna){
  clusters <- as.data.frame(table(tpm.de.wgcna$module))
  clusters <- clusters[order(-clusters[,2]),1]

  tpm.de.wgcna.reordered <- as.data.frame(matrix(nrow = 0,
                                                 ncol = ncol(tpm.de.wgcna)))
  for(i in 1:length(clusters)){
    tpm.de.wgcna.reordered <- as.data.frame(rbind(tpm.de.wgcna.reordered,
                                                  tpm.de.wgcna[tpm.de.wgcna$module == clusters[i] & tpm.de.wgcna$invert == F,],
                                                  tpm.de.wgcna[tpm.de.wgcna$module == clusters[i] & tpm.de.wgcna$invert == T,]))
  }
  return(tpm.de.wgcna.reordered)
}

get_heatmap_separators <- function(vector){
  sep <- c()
  for(i in 2:length(unique(vector))){
    sep[length(sep) + 1] <- min(which(vector == unique(vector)[i])) - 1
  }
  return(sep)
}


eigengene_refit <- function(tpm.de, mergedMEs, min_pearson_corr = 0.8){
  tpm.de.wgcna <- tpm.de
  tpm.de.wgcna$invert <- T
  tpm.de.wgcna$module <- NA
  for(i in 1:nrow(tpm.de)){
    minimum_correlation <- min_pearson_corr
    for(j in 1:ncol(mergedMEs)){
      cor <- abs(cor(t(tpm.de[i,]), mergedMEs[,j], method = "pearson"))
      if(cor > minimum_correlation){
        minimum_correlation <- cor
        tpm.de.wgcna$module[i] <- colnames(mergedMEs)[j]
      }
    }
    if(!is.na(tpm.de.wgcna$module[i])){
      if(cor(t(tpm.de[i,]), mergedMEs[,which(colnames(mergedMEs) == tpm.de.wgcna$module[i])], method = "pearson") > 0){
        tpm.de.wgcna$invert[i] <- F
      }
    }else{
      tpm.de.wgcna$invert[i] <- NA
    }

  }
  tpm.de.wgcna$module <- gsub("^ME","",tpm.de.wgcna$module)
  return(tpm.de.wgcna)
}


consolidate_functional_terms <- function(geneinfo){
  geneinfo_genes <- unique(geneinfo$gene)
  ipr <- data.frame()
  gobio <- data.frame()
  gocell <- data.frame()
  gomol <- data.frame()

  ipr.df <- as.data.frame(unlist(strsplit(paste(geneinfo$interpro_description, collapse = "|"),  split = "[|]")))
  gobio.df <- as.data.frame(unlist(strsplit(paste(geneinfo$go_biologicalprocess, collapse = "|"),  split = "[|]")))
  gocell.df <- as.data.frame(unlist(strsplit(paste(geneinfo$go_cellularcomponent, collapse = "|"),  split = "[|]")))
  gomol.df <- as.data.frame(unlist(strsplit(paste(geneinfo$go_molecularfunction, collapse = "|"),  split = "[|]")))

  ipr <- rbind(ipr, ipr.df)
  gobio <- rbind(gobio, gobio.df)
  gocell <- rbind(gocell, gocell.df)
  gomol <- rbind(gomol, gomol.df)

  ipr <- as.data.frame(table(ipr))
  gobio <- as.data.frame(table(gobio))
  gocell <- as.data.frame(table(gocell))
  gomol <- as.data.frame(table(gomol))

  functionalterms.list <- list(ipr = ipr, gobio = gobio, gocell = gocell, gomol = gomol)

  return(functionalterms.list)
}



functionaltermenrichment <- function(genes, geneinfo, functionalterms.list){

  geneinfo.subset <- geneinfo[geneinfo$gene %in% genes,]
  term <- c()
  clusteroccurences <- c()
  genomeoccurences <- c()
  pvalue <- c()
  correctedpvalue <- c()
  oddsratio <- c()
  clusterProportion <- c()
  genomeProportion <- c()
  clusterSize <- c()
  genomeSize <- c()
  newOddsRatio <- c()

  functionalterms.list.subset <- consolidate_functional_terms(geneinfo.subset)

  for(i in 1:length(functionalterms.list)){
    for(j in 1:nrow(functionalterms.list[[i]])){

      freq.all <- functionalterms.list[[i]][j,2] ## How amany times the IPR/GO term appears in the entire geneinfo file

      ## determine if the functional term is in the subset
      ## If it isn't, then the frequency is set to 0
      ## If it is in the subset, then extract the frequency of the IPR/GO term from within the subset
      freq.subset <- ifelse(functionalterms.list[[i]][j,1] %in% functionalterms.list.subset[[i]][,1],
                            functionalterms.list.subset[[i]][functionalterms.list.subset[[i]][,1] == as.character(functionalterms.list[[i]][j,1]),2],
                            0)

      ## Number of genes in the whole genome
      genes.all <- nrow(geneinfo)

      ## Number of genes in the cluster
      genes.subset <- nrow(geneinfo.subset)
      # [col.row]
      ## [1,1] is frequency is the subset, [1,2] is remaining frequency in rest of gene term file (col 1 total is total frequency of term in genome)
      ## [2,1] is no. genes that do not have this term in the subset, [2,2] is no. genes that do not have this term in genome (minus genes in subset)
      fisherexact.matrix <- matrix(c(freq.subset, freq.all - freq.subset,
                                     genes.subset - freq.subset, genes.all - genes.subset - freq.all + freq.subset),
                                   nrow = 2,
                                   ncol = 2)
      fisher.test <- fisher.test(fisherexact.matrix)

      term[length(term) + 1] <- as.character(functionalterms.list[[i]][j,1])
      clusteroccurences[length(clusteroccurences) + 1] <- as.numeric(as.character(freq.subset))
      genomeoccurences[length(genomeoccurences) + 1] <- as.numeric(as.character(freq.all))
      pvalue[length(pvalue) + 1] <- as.numeric(as.character(fisher.test$p.value))
      correctedpvalue[length(correctedpvalue) + 1] <- p.adjust(as.numeric(as.character(fisher.test$p.value)), method = "fdr", n = nrow(functionalterms.list[[i]]))
      oddsratio[length(oddsratio) + 1] <- as.numeric(as.character(fisher.test$estimate)) ## if < 1, term is under represented in the clutser

      clusterProportion[length(clusterProportion) + 1] <- as.numeric(fisherexact.matrix[1,1]/fisherexact.matrix[2,1])
      clusterSize[length(clusterSize) + 1] <- as.numeric(fisherexact.matrix[2,1])
      genomeProportion[length(genomeProportion) + 1] <- as.numeric(fisherexact.matrix[1,2]/fisherexact.matrix[2,2])
      genomeSize[length(genomeSize) + 1] <- as.numeric(fisherexact.matrix[2,2])

      newOddsRatio[length(newOddsRatio) + 1] <- as.numeric((fisherexact.matrix[1,1]/fisherexact.matrix[2,1])/(fisherexact.matrix[1,2]/fisherexact.matrix[2,2]))

    }
  }

  terms.df <- as.data.frame(cbind(term,
                                  clusteroccurences,
                                  genomeoccurences,
                                  pvalue,
                                  correctedpvalue,
                                  oddsratio,
                                  clusterProportion,
                                  clusterSize,
                                  genomeProportion,
                                  genomeSize,
                                  newOddsRatio))
  terms.df <- terms.df[order(as.numeric(as.character(terms.df$pvalue))),]
  return(terms.df)
}



variance_function <- function(counts.to.use, groups.to.use, form){
  y <- edgeR::DGEList(counts = counts.to.use)
  y <- edgeR::calcNormFactors(y)
  design <- stats::model.matrix(~groups.to.use$Infection.JirdSex.WormSex)
  
  y <- limma::voom(y, design)
  
  varPart <- fitExtractVarPartModel(y, form, groups.to.use)
  
  varPart.df <- gg_plotvar(varPart)
  
  return(varPart.df)
}

gg_plotvar <- function(varPart){
  varPart.df <- data.frame()
  for(col in colnames(varPart)){
    df <- as.data.frame(varPart) %>% select(all_of(col))
    df$attribute <- col
    colnames(df)[1] <- 'Variance_Explained'
    
    varPart.df <- rbind(varPart.df, df)
  }
  
  
  varPart.df$attribute <- factor(varPart.df$attribute, levels = unique(varPart.df$attribute))
  
  print(
    ggplot(varPart.df, aes(x = attribute, y = Variance_Explained * 100)) + geom_boxplot() + labs(y = 'Variance explained (%)')
  )
  
  return(varPart.df)

}


ggplot_legend <- function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)

}


calculate_DE_genes <- function(counts.mat, groups.df, tpm.mat){
  FDRcutoff <- 0.05
  cpm.cutoff <- 5/min(colSums(counts.mat)) * 1000000 
  print(paste0(nrow(counts)," total genes"))
  y <- edgeR::DGEList(counts = counts.mat, group = groups.df$differential_category)
  y <- edgeR::calcNormFactors(y)
  keep <- rowSums(cpm(y) >= cpm.cutoff) >= min(table(groups.df$differential_category)) 
  keep.df <- as.data.frame(table(keep))
  print(paste0(keep.df[keep.df[,1] == F,2]," genes excluded using edgeR filter"))
  y <- y[keep, , keep.lib.sizes = F] 
  design <- stats::model.matrix(~groups.df$differential_category)
  y <- estimateDisp(y , design)
  fit <- edgeR::glmQLFit(y, design) 
  qlf <- edgeR::glmQLFTest(fit, coef = 2:ncol(fit)) 
  FDR <- as.data.frame(p.adjust(qlf$table$PValue, method="BH"))
  rownames(FDR) <- rownames(qlf$table)
  colnames(FDR) <- FDR
  degenes <- rownames(FDR[FDR[,1] < FDRcutoff, , drop = F])
  print(paste0(length(degenes)," DE genes"))
  
  if(length(degenes) > 5){
    tpm.de <- tpm.mat[rownames(tpm.mat) %in% degenes,]
  }else{
    tpm.de <- tpm.mat[rownames(tpm.mat) %in% rownames(y$counts),]
  }
  
  return(tpm.de)
  
}


plot_genes <- function(tpm.de, groups.df, OUTFILE, log_transform = FALSE){
  
  tpm_dendrogram <- pvclust::pvclust(as.data.frame(t(scale(t(log2(tpm.de+1))))), method.dist="cor", method.hclust="average",
                                   nboot=1000, parallel=TRUE)
  
  pdf(paste0(FIGURE_DIR, "/dendrogram_de_zscore_", OUTFILE, ".pdf"),
      width = 20,
      height = 10)
    plot(tpm_dendrogram)
  invisible(dev.off())
  
  
  dendro.data <- ggdendro::dendro_data(tpm_dendrogram$hclust)
  dendro_order <- dendro.data$labels %>% select('label')
  
  sample_order <- factor(as.character(dendro_order$label), levels = as.character(dendro_order$label))
  
  groups.df <- factor_groups(groups.df[order(match(groups.df$Sample, as.character(sample_order))),])
  
  rownames(groups.df) <- NULL
  
  
  if(log_transform){
    log.tpm.de <- log2(tpm.de + 1)
    heatmap_matrix <- as.matrix(log.tpm.de)
    breaks <- seq(min(heatmap_matrix),max(heatmap_matrix),by=.5)
    hmcol <- colorRampPalette(c("white","firebrick1","firebrick3"))(length(breaks)-1)
  }else{
    log.tpm.de <- log2(tpm.de + 1)
    zscore.log.tpm.de <- as.data.frame(t(scale(t(log.tpm.de))))
    heatmap_matrix <- as.matrix(zscore.log.tpm.de)
    hmcol <- colorRampPalette(c("navyblue","white","firebrick3"))(12)
    breaks <- seq(-3,3,by=.5)
  }
  
  
    colcol <- as.character(groups.df$InfectionColour)
    colcol1 <- as.character(groups.df$WormColour)
    colcol2 <- as.character(groups.df$JirdColour)
    colcol3 <- cbind(colcol, colcol1, colcol2)
    colnames(colcol3) <- c('Infection', 'WormStage', 'GerbilSex')
    colcol3 <- as.matrix(colcol3)
      
  
  
  
  
  heatmap_matrix <- heatmap_matrix[,as.character(sample_order)]
  
  pdf(paste0(FIGURE_DIR, "/", OUTFILE, ".pdf"),
      width = 20,
      height = 10)
    heatmap.3(heatmap_matrix,
                  col=hmcol,
                  trace="none",
                  labRow = NA,
                  labCol = F,
                  Rowv = T,
                  Colv = F,
                  ColSideColorsSize = 4,
                  ColSideColors=colcol3,
                  lhei = c(2,8,1),
                  breaks = breaks,
                  density.info = 'histogram',
                  dendrogram = "both")
  
  invisible(dev.off())
}




```



# Set up directories

```{r Directories}
# Set up Directories
cwd <- getwd()
OUTPUT_DIR <- paste0(cwd, "/output_files_pairwise/")
FILE_DIR <- paste0(OUTPUT_DIR, "/files/")
FIGURE_DIR <- paste0(OUTPUT_DIR, '/figures/')

dir.create(OUTPUT_DIR, showWarnings = F, recursive = T)
dir.create(FILE_DIR, showWarnings = F, recursive = T)
dir.create(FIGURE_DIR, showWarnings = F, recursive = T)

```

## File URLS
```{r}
COUNTS_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/combined.final.counts"
GROUPS_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/groups.csv"
GENEINFO_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/bpahangi_gene.info"
GFF_URL <- "https://raw.githubusercontent.com/christopher-holt/bpahangi_host_sex_effect/refs/heads/main/input_files/GCA_012070555.1_ASM1207055v1_genomic.gff"
```



## Read in Files
```{r}
## Counts file
counts.orig <- readr::read_delim(COUNTS_URL, 
                          "\t", escape_double = FALSE, col_names = FALSE, 
                          trim_ws = TRUE, show_col_types = F)

## Read in groups file
groups <- readr::read_delim(GROUPS_URL, 
                          ",", escape_double = FALSE, col_names = FALSE, 
                          trim_ws = TRUE, show_col_types = F)


geneinfo <- readr::read_delim(GENEINFO_URL, delim = '\t', show_col_types = F, col_names = T)

b_pahangi_annotation <- readr::read_delim(GFF_URL, 
                            "\t", escape_double = FALSE, col_names = FALSE, 
                            trim_ws = TRUE,
                            show_col_types = F, comment = "#")

```

# Format Data Files

## Format groups file

```{r}
groups <- as.data.frame(groups)

## Male Jirds are Orange, Female Jirds are DarkGreen
groups$X2 <- NA
groups <- groups %>% mutate(X2 = ifelse(grepl("^M",groups$X1), 'orange',
                                        ifelse(grepl('^F', groups$X1), 'darkgreen','white')))

## Male Worms are orange, Female Worms are Blue, MF are white
groups$X3 <- NA
groups <- groups %>% mutate(X3 = ifelse(grepl("-males",groups$X1), 'red',
                                        ifelse(grepl('-females', groups$X1), 'blue','white')))

## IP Infections are Grey, SQ Infections are black
groups$X4 <- NA
groups <- groups %>% mutate(X4 = ifelse(grepl("IP", groups$X1), 'grey',
                                        ifelse(grepl('SQ', groups$X1), 'black', 'white')))

## Add litter to each sample
groups$X5 <- NA
groups <- groups %>% mutate(X5 = ifelse(grepl("M200_IP_Bp|F006_IP_Bp", groups$X1), 'litter_05',
                                        ifelse(grepl("F009_IP_Bp|M204_IP_Bp", groups$X1), 'litter_02',
                                               ifelse(grepl("F012_IP_Bp|M206_IP_Bp", groups$X1), 'litter_03',
                                                      ifelse(grepl("F024_SQ_Bp|M055_SQ_Bp", groups$X1), 'litter_12',
                                                             ifelse(grepl("M057_SQ_Bp", groups$X1), 'litter_13',
                                                                    ifelse(grepl("F015_SQ_Bp", groups$X1), 'litter_08',
                                                                           ifelse(grepl("M069_SQ_Bp", groups$X1), 'CRL',
                                                                                  ifelse(grepl("F022_SQ_Bp", groups$X1), 'litter_11', NA)))))))))

## Add Infection classifier to be able to select and group IP and SQ samples separately
groups$X6 <- NA
groups <- groups %>% mutate(X6 = ifelse(grepl("IP", groups$X1), "IP", "SQ")) 

## Add Host Sex classifier to be able to select and group Male Jird and Female Jird separately
groups$X7 <- NA
groups <- groups %>% mutate(X7 = ifelse(grepl("^F", groups$X1), "FemaleGerbil", "MaleGerbil"))

## Add Worm Sex classifier to be able to select and group Male Worm and Female Worm separately

groups$X8 <- NA
groups <- groups %>% mutate(X8 = ifelse(grepl("-females", groups$X1), "FemaleWorm", 
                                        ifelse(grepl("-males", groups$X1), "MaleWorm", "mf")))

## Combine Infection and Host Sex classifiers 
groups$X9 <- NA
groups$X9 <- paste0(groups$X6, "_", groups$X7)
## Combine Infection and Worm Sex Classifiers
groups$X10 <- NA
groups$X10 <- paste0(groups$X6, "_",  groups$X8)
## Combine Host Sex and Worm Sex classifiers
groups$X11 <- NA
groups$X11 <- paste0(groups$X7, "_", groups$X8)
## Combine all three classifiers together
groups$X12 <- NA
groups$X12 <- paste0(groups$X6, "_", groups$X7, "_", groups$X8)

groups$X13 <- NA
groups <- groups %>% mutate(X13 = ifelse(grepl("IP_FemaleWorm", groups$X10), 'blue',
                                             ifelse(grepl("IP_MaleWorm", groups$X10), 'red',
                                                    ifelse(grepl("SQ_FemaleWorm", groups$X10), 'lightblue',
                                                           ifelse(grepl("SQ_MaleWorm", groups$X10), 'pink',
                                                                  ifelse(grepl("IP_mf", groups$X10), 'darkgreen',
                                                                         ifelse(grepl("SQ_mf", groups$X10), 'lightgreen', NA)))))))



## Set Litter Colour
groups$X14 <- NA
groups <- groups %>% mutate(X14 = ifelse(grepl('litter_05', groups$X5), 'green',
                                        ifelse(grepl('litter_02', groups$X5), 'orange',
                                               ifelse(grepl('litter_03', groups$X5), 'blue',
                                                      ifelse(grepl('litter_12', groups$X5), 'violet',
                                                             ifelse(grepl('litter_13', groups$X5), 'black',
                                                                    ifelse(grepl('litter_08', groups$X5), 'grey',
                                                                           ifelse(grepl('CRL', groups$X5), 'white',
                                                                                  ifelse(grepl('litter_11', groups$X5), 'gold', NA)))))))))


groups$X15 <- NA
groups <- groups %>% 
  dplyr::mutate(X15 = ifelse(grepl("F015_SQ_Bp-females", groups$X1), 'blue',
                                           ifelse(grepl("F015_SQ_Bp-males", groups$X1), 'orange',
                                                 ifelse(grepl("F024_SQ_Bp-females", groups$X1), 'green',
                                                        ifelse(grepl('M069_SQ_Bp-males', groups$X1), 'black',
                                                               ifelse(grepl('M057_SQ_Bp-males', groups$X1), 'grey',
                                                                      'white'))))))
groups$X16 <- NA
groups <- groups %>% 
  dplyr::mutate(X16 = ifelse(grepl("F015_SQ_Bp-females", groups$X1), "F015_SQ_Bp-females",
                                           ifelse(grepl("F015_SQ_Bp-males", groups$X1), 'F015_SQ_Bp-males',
                                                 ifelse(grepl("F024_SQ_Bp-females", groups$X1), 'F024_SQ_Bp-females',
                                                        ifelse(grepl('M069_SQ_Bp-males', groups$X1), 'M069_SQ_Bp-males',
                                                               ifelse(grepl('M057_SQ_Bp-males', groups$X1), 'M057_SQ_Bp-males',
                                                                      'not_outlier'))))))
groups$X17 <- NA
groups <- groups %>%
  dplyr::mutate(X17 = ifelse(grepl('MaleGerbil_FemaleWorm', groups$X11), 'blue',
                             ifelse(grepl('MaleGerbil_MaleWorm', groups$X11), 'red',
                                    ifelse(grepl('MaleGerbil_mf', groups$X11), 'darkgreen',
                                           ifelse(grepl('FemaleGerbil_FemaleWorm', groups$X11), 'lightblue',
                                                  ifelse(grepl('FemaleGerbil_MaleWorm', groups$X11), 'pink',
                                                         ifelse(grepl('FemaleGerbil_mf', groups$X11), 'green', NA)))))))

colnames(groups) <- c("Sample", "JirdColour", "WormColour", 
                      "InfectionColour", "Litter", "Infection", "JirdSex",
                      "WormSex", "Infection.JirdSex", "Infection.WormSex", "JirdSex.WormSex", "Infection.JirdSex.WormSex",
                      'Infection.WormSex.Colour', 'LitterColour', 'outlier_samples', 'outliersample_names','JirdSex.WormSex.Colour')

groups <- factor_groups(groups)


```


## Format Annotation File

```{r format annotation file}
b_pahangi_annotation <- b_pahangi_annotation %>% tidyr::separate('X9', into = c("Name", "X9"), sep = ';')
b_pahangi_annotation$X9 <- gsub(".*:", "", b_pahangi_annotation$X9)
b_pahangi_annotation$X9 <- gsub(".*=", "", b_pahangi_annotation$X9)
b_pahangi_annotation$Name <- gsub(".*:", "", b_pahangi_annotation$Name)
b_pahangi_annotation$Name <- gsub(".*=", "", b_pahangi_annotation$Name)

b_pahangi_genes <- b_pahangi_annotation %>% dplyr::filter(X3 == "gene")

b_pahangi_genes$length <- b_pahangi_genes$X5 - b_pahangi_genes$X4

b_pahangi_gene_length <- b_pahangi_genes %>% dplyr::select('Name', 'length')

```

## Format counts table into a matrix

```{r format counts table into a matrix}
colnames(counts.orig) <- c("Name", "counts", "sample")

counts.orig$counts <- as.double(counts.orig$counts)
counts.orig$sample <- str_replace_all(counts.orig$sample, ".counts", "")

counts.orig <- counts.orig %>% dplyr::arrange(sample, Name)
counts.orig <- counts.orig[!grepl("__", counts.orig$Name),]

counts.orig <- dplyr::inner_join(counts.orig, b_pahangi_gene_length, by = "Name") %>% dplyr::arrange(sample, Name)
counts <- create_matrix(df = counts.orig, row = "Name", centre = "counts", column = "sample")

```


## Select genes annotated as mRNA

```{r}
b_pahangi_mRNAs <- b_pahangi_annotation %>% filter(X3 == "mRNA")

b_pahangi_mRNAs <- unique(b_pahangi_mRNAs$X9)

counts <- counts[(rownames(counts) %in% b_pahangi_mRNAs),]

## Select genes in the counts.orig table that are in the counts matrix
counts.orig <- counts.orig[(counts.orig$Name %in% rownames(counts)),]
```


## Calculate TPMs

```{r calculate TPMs}

if(all(rownames(counts) == unique(counts.orig$Name))){ ## Makes sure the genes are in the same order for the correct division
  
  tpm <- calculate_TPM(counts.mat = counts, counts.orig = counts.orig)

}else{
  message('The genes are not in the same order')
}

counts <- counts[,levels(groups$Sample)]
tpm <- tpm[,levels(groups$Sample)]

```


## Remove Microfilariae
```{r}
## Remove all microfilariae samples
groups <- groups %>% filter(!grepl('Bp-mf', groups$Sample))
groups <- factor_groups(groups)

counts <- counts[match(rownames(tpm),rownames(counts)),match(groups$Sample,colnames(counts))]
tpm <- tpm[,match(groups$Sample,colnames(tpm))]

```


## Differences in Host Sex Effect
### IP Male Worms
```{r}
groups.ip.maleworm <- groups %>% filter(Infection.WormSex == "IP_MaleWorm")
groups.ip.maleworm$differential_category <- groups.ip.maleworm$JirdSex
groups.ip.maleworm <- factor_groups(groups.ip.maleworm)

counts.ip.maleworm <- counts[,levels(groups.ip.maleworm$Sample)]
tpm.ip.maleworm <- tpm[,levels(groups.ip.maleworm$Sample)]

## 0 DE Genes
tpm.ip.maleworm.de <- calculate_DE_genes(counts.mat = counts.ip.maleworm, groups.df = groups.ip.maleworm, tpm.mat = tpm.ip.maleworm)

plot_genes(tpm.de = tpm.ip.maleworm.de, groups.df = groups.ip.maleworm, OUTFILE = "IP_MaleWorm")
plot_genes(tpm.de = tpm.ip.maleworm.de, groups.df = groups.ip.maleworm, OUTFILE = "IP_MaleWorm_log", log_transform = TRUE)
```

### IP Female Worms
```{r}
groups.ip.femaleworm <- groups %>% filter(Infection.WormSex == "IP_FemaleWorm")
groups.ip.femaleworm$differential_category <- groups.ip.femaleworm$JirdSex
groups.ip.femaleworm <- factor_groups(groups.ip.femaleworm)

counts.ip.femaleworm <- counts[,levels(groups.ip.femaleworm$Sample)]
tpm.ip.femaleworm <- tpm[,levels(groups.ip.femaleworm$Sample)]

## 0 DE Genes
tpm.ip.femaleworm.de <- calculate_DE_genes(counts.mat = counts.ip.femaleworm, groups.df = groups.ip.femaleworm, tpm.mat = tpm.ip.femaleworm)

plot_genes(tpm.de = tpm.ip.femaleworm.de, groups.df = groups.ip.femaleworm, OUTFILE = "IP_FemaleWorm")
plot_genes(tpm.de = tpm.ip.femaleworm.de, groups.df = groups.ip.femaleworm, OUTFILE = "IP_FemaleWorm_log", log_transform = TRUE)
```

### SQ Male Worms
```{r}
groups.sq.maleworm <- groups %>% filter(Infection.WormSex == "SQ_MaleWorm")
groups.sq.maleworm$differential_category <- groups.sq.maleworm$JirdSex
groups.sq.maleworm <- factor_groups(groups.sq.maleworm)

counts.sq.maleworm <- counts[,levels(groups.sq.maleworm$Sample)]
tpm.sq.maleworm <- tpm[,levels(groups.sq.maleworm$Sample)]

## 0 DE Genes
tpm.sq.maleworm.de <- calculate_DE_genes(counts.mat = counts.sq.maleworm, groups.df = groups.sq.maleworm, tpm.mat = tpm.sq.maleworm)

plot_genes(tpm.de = tpm.sq.maleworm.de, groups.df = groups.sq.maleworm, OUTFILE = "SQ_MaleWorm")
plot_genes(tpm.de = tpm.sq.maleworm.de, groups.df = groups.sq.maleworm, OUTFILE = "SQ_MaleWorm_log", log_transform = TRUE)
```

### SQ Female Worms
```{r}
groups.sq.femaleworm <- groups %>% filter(Infection.WormSex == "SQ_FemaleWorm")
groups.sq.femaleworm$differential_category <- groups.sq.femaleworm$JirdSex
groups.sq.femaleworm <- factor_groups(groups.sq.femaleworm)

counts.sq.femaleworm <- counts[,levels(groups.sq.femaleworm$Sample)]
tpm.sq.femaleworm <- tpm[,levels(groups.sq.femaleworm$Sample)]

## 0 DE Genes
tpm.sq.femaleworm.de <- calculate_DE_genes(counts.mat = counts.sq.femaleworm, groups.df = groups.sq.femaleworm, tpm.mat = tpm.sq.femaleworm)

plot_genes(tpm.de = tpm.sq.femaleworm.de, groups.df = groups.sq.femaleworm, OUTFILE = "SQ_FemaleWorm")
plot_genes(tpm.de = tpm.sq.femaleworm.de, groups.df = groups.sq.femaleworm, OUTFILE = "SQ_FemaleWorm_log", log_transform = TRUE)



```

## No effect from Host Sex
## Differences in Life Stage
### IP Worms from Male Gerbil
```{r}
groups.ip.malegerbil <- groups %>% filter(Infection.JirdSex == "IP_MaleGerbil")
groups.ip.malegerbil$differential_category <- groups.ip.malegerbil$WormSex
groups.ip.malegerbil <- factor_groups(groups.ip.malegerbil)

counts.ip.malegerbil <- counts[,levels(groups.ip.malegerbil$Sample)]
tpm.ip.malegerbil <- tpm[,levels(groups.ip.malegerbil$Sample)]

## 5462 DE Genes
tpm.ip.malegerbil.de <- calculate_DE_genes(counts.mat = counts.ip.malegerbil, groups.df = groups.ip.malegerbil, tpm.mat = tpm.ip.malegerbil)

plot_genes(tpm.de = tpm.ip.malegerbil.de, groups.df = groups.ip.malegerbil, OUTFILE = "IP_MaleGerbil")
plot_genes(tpm.de = tpm.ip.malegerbil.de, groups.df = groups.ip.malegerbil, OUTFILE = "IP_MaleGerbil_log", log_transform = TRUE)


```

### SQ Worms from Male Gerbil
```{r}
groups.sq.malegerbil <- groups %>% filter(Infection.JirdSex == "SQ_MaleGerbil")
groups.sq.malegerbil$differential_category <- groups.sq.malegerbil$WormSex
groups.sq.malegerbil <- factor_groups(groups.sq.malegerbil)

counts.sq.malegerbil <- counts[,levels(groups.sq.malegerbil$Sample)]
tpm.sq.malegerbil <- tpm[,levels(groups.sq.malegerbil$Sample)]

## 0 DE Genes
tpm.sq.malegerbil.de <- calculate_DE_genes(counts.mat = counts.sq.malegerbil, groups.df = groups.sq.malegerbil, tpm.mat = tpm.sq.malegerbil)


plot_genes(tpm.de = tpm.sq.malegerbil.de, groups.df = groups.sq.malegerbil, OUTFILE = "SQ_MaleGerbil")
plot_genes(tpm.de = tpm.sq.malegerbil.de, groups.df = groups.sq.malegerbil, OUTFILE = "SQ_MaleGerbil_log", log_transform = TRUE)

```



### IP Worms from Female Gerbil
```{r}
groups.ip.femalegerbil <- groups %>% filter(Infection.JirdSex == "IP_FemaleGerbil")
groups.ip.femalegerbil$differential_category <- groups.ip.femalegerbil$WormSex
groups.ip.femalegerbil <- factor_groups(groups.ip.femalegerbil)

counts.ip.femalegerbil <- counts[,levels(groups.ip.femalegerbil$Sample)]
tpm.ip.femalegerbil <- tpm[,levels(groups.ip.femalegerbil$Sample)]

## 7055 DE Genes
tpm.ip.femalegerbil.de <- calculate_DE_genes(counts.mat = counts.ip.femalegerbil, groups.df = groups.ip.femalegerbil, tpm.mat = tpm.ip.femalegerbil)

plot_genes(tpm.de = tpm.ip.femalegerbil.de, groups.df = groups.ip.femalegerbil, OUTFILE = "IP_FemaleGerbil")
plot_genes(tpm.de = tpm.ip.femalegerbil.de, groups.df = groups.ip.femalegerbil, OUTFILE = "IP_FemaleGerbil_log", log_transform = TRUE)


```




### SQ Worms from Female Gerbil
```{r}
groups.sq.femalegerbil <- groups %>% filter(Infection.JirdSex == "SQ_FemaleGerbil")
groups.sq.femalegerbil$differential_category <- groups.sq.femalegerbil$WormSex
groups.sq.femalegerbil <- factor_groups(groups.sq.femalegerbil)

counts.sq.femalegerbil <- counts[,levels(groups.sq.femalegerbil$Sample)]
tpm.sq.femalegerbil <- tpm[,levels(groups.sq.femalegerbil$Sample)]

## 0
tpm.sq.femalegerbil.de <- calculate_DE_genes(counts.mat = counts.sq.femalegerbil, groups.df = groups.sq.femalegerbil, tpm.mat = tpm.sq.femalegerbil)

plot_genes(tpm.de = tpm.sq.femalegerbil.de, groups.df = groups.sq.femalegerbil, OUTFILE = "SQ_FemaleGerbil")
plot_genes(tpm.de = tpm.sq.femalegerbil.de, groups.df = groups.sq.femalegerbil, OUTFILE = "SQ_FemaleGerbil_log", log_transform = TRUE)


```



## Differences in Infection Mode
### Male Gerbils, Male Worms
```{r}
groups.malegerbil.maleworm <- groups %>% filter(JirdSex.WormSex == "MaleGerbil_MaleWorm")
groups.malegerbil.maleworm$differential_category <- groups.malegerbil.maleworm$Infection
groups.malegerbil.maleworm <- factor_groups(groups.malegerbil.maleworm)

counts.malegerbil.maleworm <- counts[,levels(groups.malegerbil.maleworm$Sample)]
tpm.malegerbil.maleworm <- tpm[,levels(groups.malegerbil.maleworm$Sample)]

## 5 DE Genes
tpm.malegerbil.maleworm.de <- calculate_DE_genes(counts.mat = counts.malegerbil.maleworm, groups.df = groups.malegerbil.maleworm, tpm.mat = tpm.malegerbil.maleworm)

plot_genes(tpm.de = tpm.malegerbil.maleworm.de, groups.df = groups.malegerbil.maleworm, OUTFILE = "MaleGerbil_MaleWorm_new")
plot_genes(tpm.de = tpm.sq.femalegerbil.de, groups.df = groups.sq.femalegerbil, OUTFILE = "SQ_FemaleGerbil_log", log_transform = TRUE)


```

### Male Gerbils, Female Worms
```{r}
groups.malegerbil.femaleworm <- groups %>% filter(JirdSex.WormSex == "MaleGerbil_FemaleWorm")
groups.malegerbil.femaleworm$differential_category <- groups.malegerbil.femaleworm$Infection
groups.malegerbil.femaleworm <- factor_groups(groups.malegerbil.femaleworm)

counts.malegerbil.femaleworm <- counts[,levels(groups.malegerbil.femaleworm$Sample)]
tpm.malegerbil.femaleworm <- tpm[,levels(groups.malegerbil.femaleworm$Sample)]

## 0 DE Genes
tpm.malegerbil.femaleworm.de <- calculate_DE_genes(counts.mat = counts.malegerbil.femaleworm, groups.df = groups.malegerbil.femaleworm, tpm.mat = tpm.malegerbil.femaleworm)

plot_genes(tpm.de = tpm.malegerbil.femaleworm.de, groups.df = groups.malegerbil.femaleworm, OUTFILE = "MaleGerbil_FemaleWorm")
plot_genes(tpm.de = tpm.sq.femalegerbil.de, groups.df = groups.sq.femalegerbil, OUTFILE = "SQ_FemaleGerbil_log", log_transform = TRUE)


```

### Female Gerbils, Male Worms
```{r}
groups.femalegerbil.maleworm <- groups %>% filter(JirdSex.WormSex == "FemaleGerbil_MaleWorm")
groups.femalegerbil.maleworm$differential_category <- groups.femalegerbil.maleworm$Infection
groups.femalegerbil.maleworm <- factor_groups(groups.femalegerbil.maleworm)

counts.femalegerbil.maleworm <- counts[,levels(groups.femalegerbil.maleworm$Sample)]
tpm.femalegerbil.maleworm <- tpm[,levels(groups.femalegerbil.maleworm$Sample)]

## 0 DE Genes
tpm.femalegerbil.maleworm.de <- calculate_DE_genes(counts.mat = counts.femalegerbil.maleworm, groups.df = groups.femalegerbil.maleworm, tpm.mat = tpm.femalegerbil.maleworm)

plot_genes(tpm.de = tpm.femalegerbil.maleworm.de, groups.df = groups.femalegerbil.maleworm, OUTFILE = "FemaleGerbil_MaleWorm")
plot_genes(tpm.de = tpm.sq.femalegerbil.de, groups.df = groups.sq.femalegerbil, OUTFILE = "SQ_FemaleGerbil_log", log_transform = TRUE)


```


### Female Gerbils, Female Worms
```{r}
groups.femalegerbil.femaleworm <- groups %>% filter(JirdSex.WormSex == "FemaleGerbil_FemaleWorm")
groups.femalegerbil.femaleworm$differential_category <- groups.femalegerbil.femaleworm$Infection
groups.femalegerbil.femaleworm <- factor_groups(groups.femalegerbil.femaleworm)

counts.femalegerbil.femaleworm <- counts[,levels(groups.femalegerbil.femaleworm$Sample)]
tpm.femalegerbil.femaleworm <- tpm[,levels(groups.femalegerbil.femaleworm$Sample)]

## 0 DE Genes
tpm.femalegerbil.femaleworm.de <- calculate_DE_genes(counts.mat = counts.femalegerbil.femaleworm, groups.df = groups.femalegerbil.femaleworm, tpm.mat = tpm.femalegerbil.femaleworm)

plot_genes(tpm.de = tpm.femalegerbil.femaleworm.de, groups.df = groups.femalegerbil.femaleworm, OUTFILE = "FemaleGerbil_FemaleWorm")
plot_genes(tpm.de = tpm.sq.femalegerbil.de, groups.df = groups.sq.femalegerbil, OUTFILE = "SQ_FemaleGerbil_log", log_transform = TRUE)


```





